---
title: "stanEx2"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show 
#date: "`r Sys.Date()`"

---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=T,warning=F,message=F)
suppressWarnings(
  suppressMessages(
    suppressPackageStartupMessages({
      library(stats)
      library(MASS)
      library(tidyverse)
    })
  )
)
```

```{r}
options(scipen=100,digits=3)
library(cmdstanr)
options(mc.cores=parallel::detectCores())
options(cmdstanr_max_rows=1000)
library(gridExtra)
```


execute mcmc sampling
```{r}
goMCMC=function(mdl,data,smp=500,wrm=100,th=1){
  mcmc=mdl$sample(
  data=data,
  seed=1,
  chains=4,
  iter_sampling=smp,
  iter_warmup=wrm,
  thin=th,
  refresh=1000
  )
  mcmc
}
```

see mcmc result and parameters
```{r}
seeMCMC=function(mcmc,exc='',ch=T){ # exclude 'exc' parameters from seeing
  print(mcmc)
  prs=mcmc$metadata()$model_params[-1] # reject lp__
  for(pr in prs){
    if(grepl('^y',pr)) next # not show predictive value "y*" information
    if(exc!='' && grepl(paste0('^',exc),pr)) next
    drw=mcmc$draws(pr)
    if(ch){
      par(mfrow=c(1,4))
      drw[,1,] |> plot(type='l',main='chain1',ylab=pr)
      drw[,2,] |> plot(type='l',main='chain2',ylab=pr)
      drw[,3,] |> plot(type='l',main='chain3',ylab=pr)
      drw[,4,] |> plot(type='l',main='chain4',ylab=pr)
    }

    par(mfrow=c(1,2))
    drw |> hist(main=pr,xlab='')
    drw |> density() |> plot(main=pr)    
  }
}
```


# regression

## normal linear model

### ex5-1.stan
normal regression
```
data{
  int N;
  int K;
  vector[N] y;
  matrix[N,K] X;
}
parameters{
  vector[K] b;
  real<lower=0> s;
}
model{
  vector[N] m=X*b;
  y~normal(m,s);
}
generated quantities{
  vector[N] y1;
  vector[N] m1=X*b;
  for(i in 1:N){
    y1[i]=normal_rng(m1[i],s);
  }
}
```

```{r}
n=30
mdl=cmdstan_model('./ex5-1.stan')
```


### single regression
```{r}
tb=tibble(x=runif(n,0,9),
          y=rnorm(n,x,1))
f=formula(y~x)
par(mfrow=c(1,1))
plot(tb$x,tb$y)
qplot(data=tb,x,y)

X=model.matrix(f,tb)

data=list(N=nrow(X),K=ncol(X),y=tb$y,X=X)

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,exc='m',ch=F)

y1=mcmc$draws('y1')
smy=summary(y1)

tb$y1=smy$median
tb$l5=smy$q5
tb$u5=smy$q95

par(mfrow=c(1,1))
plot(tb$y,tb$y1,xlab='obs.',ylab='prd.')
abline(0,1)
qplot(data=tb,y,y1,xlab='obs.',ylab='prd.')+geom_abline(intercept=0,slope=1)

par(mfrow=c(1,2))
hist(tb$y-tb$y1,xlab='obs.-prd.',main='residual')
density(tb$y-tb$y1) |> plot(xlab='obs.-prd.',main='residual')

tb=arrange(tb,y1)

par(mfrow=c(1,1))
ylim=c(min(tb$l5),max(tb$u5))
plot(tb$y,ylim=ylim,ylab='y',col='red')
par(new=T)
plot(tb$y1,ylim=ylim,ylab='',col='black')
par(new=T)
plot(tb$l5,ylim=ylim,ylab='',type='l',col='gray')
par(new=T)
plot(tb$u5,ylim=ylim,ylab='',type='l',col='gray')



qplot(1:n,y,data=tb,col=I('red'))+
  geom_point(aes(x=1:n,y=y1),tb,col='black')+
  geom_line(aes(x=1:n,y=l5),tb,col='gray')+
  geom_line(aes(x=1:n,y=u5),tb,col='gray')
```


### multiple regression
```{r}
tb=tibble(x1=runif(n,0,9),x2=runif(n,0,9),
          y=rnorm(n,x1-x2,1))
f=formula(y~x1+x2)
par(mfrow=c(1,2))
plot(tb$x1,tb$y)
plot(tb$x2,tb$y)

X=model.matrix(f,tb)
data=list(N=nrow(X),K=ncol(X),y=tb$y,X=X)

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,exc='m',ch=F)

y1=mcmc$draws('y1')
smy=summary(y1)

tb$y1=smy$median
tb$l5=smy$q5
tb$u5=smy$q95

par(mfrow=c(1,1))
plot(tb$y,tb$y1,xlab='obs.',ylab='prd.')
abline(0,1)
qplot(data=tb,y,y1,xlab='obs.',ylab='prd.')+geom_abline(intercept=0,slope=1)

par(mfrow=c(1,2))
hist(tb$y-tb$y1,xlab='obs.-prd.',main='residual')
density(tb$y-tb$y1) |> plot(xlab='obs.-prd.',main='residual')

tb=arrange(tb,y1)
par(mfrow=c(1,1))
ylim=c(min(tb$l5),max(tb$u5))
plot(tb$y,ylim=ylim,ylab='y',col='red')
par(new=T)
plot(tb$y1,ylim=ylim,ylab='',col='black')
par(new=T)
plot(tb$l5,ylim=ylim,ylab='',type='l',col='gray')
par(new=T)
plot(tb$u5,ylim=ylim,ylab='',type='l',col='gray')



qplot(1:n,y,data=tb,col=I('red'))+
  geom_point(aes(x=1:n,y=y1),tb,col='black')+
  geom_line(aes(x=1:n,y=l5),tb,col='gray')+
  geom_line(aes(x=1:n,y=u5),tb,col='gray')
```


### ANOVA
```{r}
tb=tibble(c=sample(c('a','b','c'),n,replace=T),
          y=rnorm(n,(c=='b')*2-(c=='c')*2,1))
f=formula(y~c)
par(mfrow=c(1,1))
boxplot(y~c,tb)
qplot(data=tb,c,y,geom='boxplot')

X=model.matrix(f,tb)
data=list(N=nrow(X),K=ncol(X),y=tb$y,X=X)

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,exc='m',ch=F)

y1=mcmc$draws('y1')
smy=summary(y1)

tb$y1=smy$median
tb$l5=smy$q5
tb$u5=smy$q95

lv=c(0,1,2)
par(mfrow=c(1,1))
plot(tb$y,tb$y1,pch=lv[factor(tb$c)],xlab='obs.',ylab='prd.')
qplot(data=tb,y,y1,shape=c,size=I(2),xlab='obs.',ylab='prd.')
plot(tb$y,tb$y1,col=1+lv[factor(tb$c)],xlab='obs.',ylab='prd.')
qplot(data=tb,y,y1,col=c,xlab='obs.',ylab='prd.')

par(mfrow=c(1,2))
hist(tb$y-tb$y1,xlab='obs.-prd.',main='residual')
density(tb$y-tb$y1) |> plot(xlab='obs.-prd.',main='residual')

tb=arrange(tb,y1)
par(mfrow=c(1,1))
ylim=c(min(tb$l5),max(tb$u5))
plot(tb$y,ylim=ylim,ylab='y',col='red',pch=lv[factor(tb$c)])
par(new=T)
plot(tb$y1,ylim=ylim,ylab='',col='black',pch=lv[factor(tb$c)])
par(new=T)
plot(tb$l5,ylim=ylim,ylab='',type='l',col='gray')
par(new=T)
plot(tb$u5,ylim=ylim,ylab='',type='l',col='gray')


qplot(data=tb,1:n,y,col=c)+
  geom_point(aes(x=1:n,y=y1),tb,col='black')+
  geom_line(aes(x=1:n,y=l5),tb,col='gray')+
  geom_line(aes(x=1:n,y=u5),tb,col='gray')
```


### ANCOVA
```{r}
tb=tibble(x=runif(n,0,9),c=sample(c('a','b','c'),n,replace=T),
          y=rnorm(n,x+(c=='b')*2-(c=='c')*2,1))

f=formula(y~x+c)
par(mfrow=c(1,1))
#plot(tb$x1,tb$y,pch=tb$c1)

lv=c(0,1,2)
plot(tb$x,tb$y,pch=lv[factor(tb$c)])
qplot(data=tb,x,y,shape=c,size=I(2))
plot(tb$x,tb$y,col=1+lv[factor(tb$c)])
qplot(data=tb,x,y,col=c)

X=model.matrix(f,tb)
data=list(N=nrow(X),K=ncol(X),y=tb$y,X=X)

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,exc='m',ch=F)

y1=mcmc$draws('y1')
smy=summary(y1)

tb$y1=smy$median
tb$l5=smy$q5
tb$u5=smy$q95

lv=c(0,1,2)
par(mfrow=c(1,1))
plot(tb$y,tb$y1,pch=lv[factor(tb$c)],xlab='obs.',ylab='prd.')
abline(0,1)
plot(tb$y,tb$y1,col=1+lv[factor(tb$c)],xlab='obs.',ylab='prd.')
abline(0,1)
qplot(data=tb,y,y1,col=c,xlab='obs.',ylab='prd.')+
  geom_abline(intercept=0,slope=1)

par(mfrow=c(1,2))
hist(tb$y-tb$y1,xlab='obs.-prd.',main='residual')
density(tb$y-tb$y1) |> plot(xlab='obs.-prd.',main='residual')

tb=arrange(tb,y1)
par(mfrow=c(1,1))
ylim=c(min(tb$l5),max(tb$u5))
plot(tb$y,ylim=ylim,ylab='y',col='red',pch=lv[factor(tb$c)])
par(new=T)
plot(tb$y1,ylim=ylim,ylab='',col='black',pch=lv[factor(tb$c)])
par(new=T)
plot(tb$l5,ylim=ylim,ylab='',type='l',col='gray')
par(new=T)
plot(tb$u5,ylim=ylim,ylab='',type='l',col='gray')


qplot(data=tb,1:n,y,col=c)+
  geom_point(aes(x=1:n,y=y1),tb,col='black')+
  geom_line(aes(x=1:n,y=l5),tb,col='gray')+
  geom_line(aes(x=1:n,y=u5),tb,col='gray')
```



## interaction of variable
```{r}
n=50
mdl=cmdstan_model('./ex5-1.stan')

tb=tibble(x=runif(n,-3,3),
          ca=sample(c('a','b'),n,replace=T),cb=sample(c('a','b'),n,replace=T),
          y=rnorm(n,x+(ca=='b')+(cb=='b')-(ca=='b')*(cb=='b'),1))

grid.arrange(qplot(data=tb,x,y,col=ca),
             qplot(data=tb,x,y,col=cb),ncol=2)

fn=function(f){
X=model.matrix(f,tb)
data=list(N=nrow(X),K=ncol(X),y=tb$y,X=X)

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,exc='m',ch=F)

y1=mcmc$draws('y1')
smy=summary(y1)

tb$y1=smy$median
tb$l5=smy$q5
tb$u5=smy$q95

grid.arrange(
  qplot(data=tb,y,y1,xlab='obs.',ylab='prd.')+geom_abline(intercept=0,slope=1))

par(mfrow=c(1,2))
hist(tb$y-tb$y1,xlab='obs.-prd.',main='residual')
density(tb$y-tb$y1) |> plot(xlab='obs.-prd.',main='residual')

tb=arrange(tb,y1)
grid.arrange(
  qplot(data=tb,1:n,y,col=ca)+
    geom_point(aes(x=1:n,y=y1),tb,col='black')+
    geom_line(aes(x=1:n,y=l5),tb,col='gray')+
    geom_line(aes(x=1:n,y=u5),tb,col='gray'),
  qplot(data=tb,1:n,y,col=cb)+
    geom_point(aes(x=1:n,y=y1),tb,col='black')+
    geom_line(aes(x=1:n,y=l5),tb,col='gray')+
    geom_line(aes(x=1:n,y=u5),tb,col='gray'),
  ncol=2
)
}


f0=formula(y~x+ca)
f1=formula(y~x+ca+cb)
f2=formula(y~x+ca*cb)

fn(f0)
fn(f1)
fn(f2)
```


```{r}
tb=tibble(xa=runif(n,-2,2),xb=runif(n,-2,2),
          ca=sample(c('a','b'),n,replace=T),cb=sample(c('a','b'),n,replace=T),
          y=rnorm(n,xa+xb-xa*xb+(ca=='b')*2+(cb=='b')*2-(ca=='b')*(cb=='b'),1))

grid.arrange(qplot(data=tb,xa,y,col=ca),
             qplot(data=tb,xa,y,col=cb),
             qplot(data=tb,xb,y,col=ca),
             qplot(data=tb,xb,y,col=cb))


fn=function(f){
X=model.matrix(f,tb)
data=list(N=nrow(X),K=ncol(X),y=tb$y,X=X)

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,exc='m',ch=F)

y1=mcmc$draws('y1')
smy=summary(y1)

tb$y1=smy$median
tb$l5=smy$q5
tb$u5=smy$q95

grid.arrange(
  qplot(data=tb,y,y1,xlab='obs.',ylab='prd.')+geom_abline(intercept=0,slope=1))

par(mfrow=c(1,2))
hist(tb$y-tb$y1,xlab='obs.-prd.',main='residual')
density(tb$y-tb$y1) |> plot(xlab='obs.-prd.',main='residual')

tb=arrange(tb,y1)
grid.arrange(
  qplot(data=tb,1:n,y,col=ca)+
    geom_point(aes(x=1:n,y=y1),tb,col='black')+
    geom_line(aes(x=1:n,y=l5),tb,col='gray')+
    geom_line(aes(x=1:n,y=u5),tb,col='gray'),
  qplot(data=tb,1:n,y,col=cb)+
    geom_point(aes(x=1:n,y=y1),tb,col='black')+
    geom_line(aes(x=1:n,y=l5),tb,col='gray')+
    geom_line(aes(x=1:n,y=u5),tb,col='gray'),
  ncol=2
)
}


f0=formula(y~xa+xb+ca+cb)
f1=formula(y~xa+xb+ca*cb)
f2=formula(y~xa*xb+ca*cb)

fn(f0)
fn(f1)
fn(f2)
```



## generalized linear regression

### log normal regression
```
objective variable [0,Infinity)

# of samples is N,  
log mi=sum(bj*xji),j=[0,K],i=[1,N]  
log yi~N(mi,s)  
```

### ex5-2.stan
log normal regression
```
data{
  int N;
  int K;
  vector[N] y;
  matrix[N,K] X;
}
parameters{
  vector[K] b;
  real<lower=0> s;
}
model{
  vector[N] m=X*b;
  y~lognormal(m,s);
}
generated quantities{
  vector[N] y1;
  vector[N] m1=X*b;
  for(i in 1:N){
    y1[i]=lognormal_rng(m1[i],s);
  }
}
```

```{r}
n=20
tb=tibble(x1=runif(n,0,9),x2=runif(n,0,9),
          y=rnorm(n,log(x1+x2),1))
f=formula(y~x1+x2)
par(mfrow=c(1,2))
plot(tb$x1,tb$y)
plot(tb$x2,tb$y)

X=model.matrix(f,tb)
data=list(N=nrow(X),K=ncol(X),y=tb$y,X=X)

mdl=cmdstan_model('./ex5-2.stan')

mle=mdl$optimize(data=data)  # it sometimes occur error and stop process
mle


mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,exc='m',ch=F)

y1=mcmc$draws('y1')
smy=summary(y1)

tb$y1=smy$median
tb$l5=smy$q5
tb$u5=smy$q95

par(mfrow=c(1,1))
plot(tb$y,tb$y1,xlab='obs.',ylab='prd.')
abline(0,1)
qplot(data=tb,y,y1,xlab='obs.',ylab='prd.')+geom_abline(intercept=0,slope=1)

par(mfrow=c(1,2))
hist(tb$y-tb$y1,xlab='obs.-prd.',main='residual')
density(tb$y-tb$y1) |> plot(xlab='obs.-prd.',main='residual')

tb=arrange(tb,y1)
par(mfrow=c(1,1))
ylim=c(min(tb$l5),max(tb$u5))
plot(tb$y,ylim=ylim,ylab='y',col='red')
par(new=T)
plot(tb$y1,ylim=ylim,ylab='',col='black')
par(new=T)
plot(tb$l5,ylim=ylim,ylab='',type='l',col='gray')
par(new=T)
plot(tb$u5,ylim=ylim,ylab='',type='l',col='gray')



qplot(1:n,y,data=tb,col=I('red'))+
  geom_point(aes(x=1:n,y=y1),tb,col='black')+
  geom_line(aes(x=1:n,y=l5),tb,col='gray')+
  geom_line(aes(x=1:n,y=u5),tb,col='gray')
```



### poisson regression
```
objective variable [0,Infinity), integer. variance of error is near to mean  
(normal linear regression, correlation is near to 1,-1, variance of error is constant)  

# of samples is N,  
log li=sum(bj*xji),j=[0,K],i=[1,N]  
yi~Po(li)  
 or  
li=sum(bj*xji),j=[0,k]  
yi~Po(exp li)  

when xj -> xj +1, y -> y* exp bj   
```

### ex6-1.stan
poisson regression
```
data{
  int N;
  int K;
  array[N] int y;
  matrix[N,K] X;
}
parameters{
  vector[K] b;
}
model{
  vector[N] l=X*b;
  y~poisson_log(l);
}
generated quantities{
  array[N] int y1;
  vector[N] l1=X*b;
  for(i in 1:N){
    y1[i]=poisson_log_rng(l1[i]);
  }
}
```

```{r}
n=30
tb=tibble(x=runif(n,-1,1),c=sample(c('a','b'),n,replace=T),
          y=rpois(n,exp(x+(c=='b')*0.5)))
f=formula(y~x+c)
qplot(data=tb,x,y,col=c)

glm(f,tb,family='poisson')

X=model.matrix(f,tb)

data=list(N=nrow(X),K=ncol(X),y=tb$y,X=X)

mdl=cmdstan_model('./ex6-1.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,exc='l1',ch=F)

y1=mcmc$draws('y1')
smy=summary(y1)

table(tb$y,smy$median)
cat('\n')
table(tb$y,smy$median,tb$c)

par(mfrow=c(1,2))
hist(tb$y-smy$median,xlab='obs.-prd.',main='residual')
density(tb$y-smy$median) |> plot(xlab='obs.-prd.',main='residual')


map=c()
for(i in 1:n){
  a=table(y1[,,i])
  map[i]=as.integer(names(a[a==max(a)]))
}
table(tb$y,map)
cat('\n')
table(tb$y,map,tb$c)

par(mfrow=c(1,2))
hist(tb$y-map,xlab='obs.-map',main='residual')
density(tb$y-map) |> plot(xlab='obs.-map',main='residual')
```


### logistic regression   
```
# of samples is N,  
objective variable 0/1 binary  
  
probability of incident pi[0,1]  
logit pi=log(pi/ 1-pi)=sum(bj*xji),j=[0,K],i=[1,N] (-Infinity, Infinity)  

yi~Ber(pi), 0/1 binary  

odds(x)=p(x)/ 1-p(x), probablity of incident / probablity of no incident  
odds ratio(x0,x1)=odds(x1)/odd(x0)  
  
when xj -> xj +1, odds ratio -> odds ratio *exp bj  
```

### ex6-2.stan
logistic regression 
```
data{
  int N;
  int K;
  array[N] int y;
  matrix[N,K] X;
}
parameters{
  vector[K] b;
}
model{
  vector[N] z=X*b;
  y~bernoulli_logit(z);
}
generated quantities{
  array[N] int y1;
  vector[N] z1=X*b;
  for(i in 1:N){
    y1[i]=bernoulli_rng(inv_logit(z1[i]));
  }
}
```

```{r}
n=30
x=runif(n,-1,1)
c=sample(c('a','b'),n,replace=T)
z=x+(c=='b')
y=rbinom(n,1,1/(1+exp(-z)))
tb=tibble(x=x,c=c,y=y)

f=formula(y~x+c)
qplot(data=tb,x,y,col=c)

glm(f,tb,family='binomial') # it can caluculte when all trials are once

X=model.matrix(f,tb)
data=list(N=nrow(X),K=ncol(X),y=tb$y,X=X)

mdl=cmdstan_model('./ex6-2.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,exc='z1',ch=F)

y1=mcmc$draws('y1')
smy=summary(y1)

table(tb$y,smy$median)
cat('\n')
table(tb$y,smy$median,tb$c)

par(mfrow=c(1,2))
hist(tb$y-smy$median,xlab='obs.-prd.',main='residual')
density(tb$y-smy$median) |> plot(xlab='obs.-prd.',main='residual')

map=c()
for(i in 1:n){
  a=table(y1[,,i])
  map[i]=as.integer(names(a[a==max(a)]))
}
table(tb$y,map)
cat('\n')
table(tb$y,map,tb$c)

par(mfrow=c(1,2))
hist(tb$y-map,xlab='obs.-map',main='residual')
density(tb$y-map) |> plot(xlab='obs.-map',main='residual')
```


### multi logistic regression  
```
# of samples is N,  
# of trials of a sample i is mi,  
objective variable [0,n], integer  
  
probability of incident pi[0,1]  
logit pi=log(pi/ 1-pi)=sum(bj*xji),j=[0,K],i=[1,N] (-Infinity, Infinity)  

yi~B(mi, pi), # of acutual incident  

odds(x)=p(x)/ 1-p(x), probablity of incident / probablity of no incident  
odds ratio(x0,x1)=odds(x1)/odd(x0)  
  
when xj -> xj +1, odds ratio -> odds ratio *exp bj  
```

### ex6-3.stan
multi logistic regression 
```
data{
  int N;
  int K;
  array[N] int m;
  array[N] int y;
  matrix[N,K] X;
}
parameters{
  vector[K] b;
}
model{
  vector[N] z=X*b;
  y~binomial_logit(m,z);
}
generated quantities{
  array[N] int y1;
  vector[N] z1=X*b;
  for(i in 1:N){
    y1[i]=binomial_rng(m[i],inv_logit(z1[i]));
  }
}
```

```{r}
n=30
m=floor(runif(n,1,10)) # trials are varying (1,10)
x=runif(n,-1,1)
c=sample(c('a','b'),n,replace=T)
z=x+(c=='b')
y=rbinom(n,m,1/(1+exp(-z)))
tb=tibble(x=x,c=c,y=y)

f=formula(y~x+c)
qplot(data=tb,x,y,col=c)

X=model.matrix(f,tb)
data=list(N=nrow(X),K=ncol(X),y=tb$y,X=X,m=m)

mdl=cmdstan_model('./ex6-3.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,exc='z1',ch=F)

y1=mcmc$draws('y1')
smy=summary(y1)

table(tb$y,smy$median)
cat('\n')
table(tb$y,smy$median,tb$c)

par(mfrow=c(1,2))
hist(tb$y-smy$median,xlab='obs.-prd.',main='residual')
density(tb$y-smy$median) |> plot(xlab='obs.-prd.',main='residual')

map=c()
for(i in 1:n){
  a=table(y1[,,i])
  map[i]=as.integer(names(a[a==max(a)]))
}
table(tb$y,map)
cat('\n')
table(tb$y,map,tb$c)

par(mfrow=c(1,2))
hist(tb$y-map,xlab='obs.-map',main='residual')
density(tb$y-map) |> plot(xlab='obs.-map',main='residual')
```


### gamma regression
```
objective variable [0,Infinity)

# of samples is N,  
log (a/li)=sum(bj*xji),j=[0,K],i=[1,N]
li=a/exp(sum(bj*xji))
yi~Ga(a,li)  
```

### ex6-4.stan
gamma regression
```
data{
  int N;
  int K;
  vector[N] y;
  matrix[N,K] X;
}
parameters{
  vector[K] b;
  real<lower=0> a;
}
model{
  vector[N] l;
  for(i in 1:N){
    l[i]=a/exp(X[i]*b);
  }
  y~gamma(a,l);
}
```

```{r}
n=20
tb=tibble(x1=runif(n,0,2),x2=runif(n,0,2),
          y=rgamma(n,3,3/exp(x1+x2)))

f=formula(y~x1+x2)
par(mfrow=c(1,2))
plot(tb$x1,tb$y)
plot(tb$x2,tb$y)

X=model.matrix(f,tb)
data=list(N=nrow(X),K=ncol(X),y=tb$y,X=X)

mdl=cmdstan_model('./ex6-4.stan')

mle=mdl$optimize(data=data)  # it sometimes occur error and stop process
mle


mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,ch=F)
```


### negative binomial regression
```
The event with probability p do not occur y times till the event occur a times  
(negative binomial distribution has larger variance than poisson distribution)

y~NB(a,p), log(p)=X*b

y~NB(a,l0/(1+l0)) when y~Po(l), l~Ga(a,l0), l0=a/E[l]
```

### ex6-5.stan
negative binomial regression
```
data{
  int N;
  int K;
  array[N] int y;
  matrix[N,K] X;
}
parameters{
  vector[K] b;
  real<lower=0> a;
}
model{
  a~cauchy(0,5);
  y~neg_binomial_2_log(X*b,a);
}
```

```{r}
n=20
tb=tibble(x1=runif(n,-1,0),x2=runif(n,-1,0),
          y=rnbinom(n,3,exp(x1+x2)))
f=formula(y~x1+x2)
par(mfrow=c(1,2))
plot(tb$x1,tb$y)
plot(tb$x2,tb$y)

X=model.matrix(f,tb)

data=list(N=nrow(X),K=ncol(X),y=tb$y,X=X)

mdl=cmdstan_model('./ex6-5.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,ch=F)
```



### beta regression
```
using prior of binomial distribution parameter p[0,1]
y~B(n,p), p~beta(a,b)

m=E[p]=a/(a+b)
s^2=V[P]=ab/(a+b)^2/(a+b+1)

m=x*be
a=ms=inv_logit(m)*s
b=(1-m)s=(1-inv_logy(m))*s
```

### ex6-6.stan
beta regression
```
data{
  int N;
  int K;
  vector[N] p;
  matrix[N,K] X;
}
parameters{
  vector[K] be;
  real<lower=0> s;
}
model{
  vector[N] a;
  vector[N] b;
  for(i in 1:N){
    a[i]=inv_logit(X[i]*be)*s;
    b[i]=(1-inv_logit(X[i]*be))*s;
  }
  p~beta(a,b);
}
```

```{r}
n=20
tb=tibble(x1=runif(n,0,0.5),x2=runif(n,0,0.5),
          p=rbeta(n,(x1+x2)*3,(1-(x1+x2))*3))
f=formula(p~x1+x2)
par(mfrow=c(1,2))
plot(tb$x1,tb$p)
plot(tb$x2,tb$p)

X=model.matrix(f,tb)

data=list(N=nrow(X),K=ncol(X),p=tb$p,X=X)

mdl=cmdstan_model('./ex6-6.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,ch=F)
```


### beta binomial regression
```
fitting to distribution has larger variance than binomial distribution
y~betaB(n,a,b) when y~B(n,p), p~beta(a,b)

m=E[p]=a/(a+b)
s^2=V[P]=ab/(a+b)^2/(a+b+1)

m=x*be
a=ms=inv_logit(m)*s
b=(1-m)s=(1-inv_logy(m))*s
```
### ex6-7.stan
beta binomial regression
```
data{
  int N;
  int K;
  array[N] int y;
  array[N] int n;
  matrix[N,K] X;
}
parameters{
  vector[K] be;
  real<lower=0> s;
}
model{
  vector[N] a;
  vector[N] b;
  for(i in 1:N){
    a[i]=inv_logit(X[i]*be)*s;
    b[i]=(1-inv_logit(X[i]*be))*s;
  }
  y~beta_binomial(n,a,b);
  s~cauchy(0,5);
}
```

```{r}
n=20
tb=tibble(x1=runif(n,0,0.5),x2=runif(n,0,0.5),
          p=rbeta(n,(x1+x2)*3,(1-(x1+x2))*3),
          n1=floor(runif(n,5,9)),
          y=rbinom(n,n1,p))
f=formula(p~x1+x2)
par(mfrow=c(1,2))
plot(tb$x1,tb$y)
plot(tb$x2,tb$y)

X=model.matrix(f,tb)

data=list(N=nrow(X),K=ncol(X),n=tb$n1,y=tb$y,X=X)

mdl=cmdstan_model('./ex6-7.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,ch=F)
```
