---
title: "stanEx1"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show 
#date: "`r Sys.Date()`"

---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=T,warning=F,message=F)
suppressWarnings(
  suppressMessages(
    suppressPackageStartupMessages({
      library(stats)
      library(MASS)
      library(tidyverse)
    })
  )
)
```

```{r}
options(scipen=100,digits=3)
library(cmdstanr)
options(mc.cores=parallel::detectCores())
options(cmdstanr_max_rows=1000)
library(gridExtra)
```

execute mcmc sampling
```{r}
goMCMC=function(mdl,data,smp=500,wrm=100,th=1){
  mcmc=mdl$sample(
  data=data,
  seed=1,
  chains=4,
  iter_sampling=smp,
  iter_warmup=wrm,
  thin=th,
  refresh=1000
  )
  mcmc
}
```

see mcmc result and parameters
```{r}
seeMCMC=function(mcmc,exc='',ch=T){ # exclude 'exc' parameters from seeing
  print(mcmc)
  prs=mcmc$metadata()$model_params[-1] # reject lp__
  for(pr in prs){
    if(grepl('^y',pr)) next # not show predictive value "y*" information
    if(exc!='' && grepl(paste0('^',exc),pr)) next
    drw=mcmc$draws(pr)
    if(ch){
      par(mfrow=c(4,1),mar=c(1,5,1,1))
      drw[,1,] |> plot(type='l',xlab='',ylab=pr)
      drw[,2,] |> plot(type='l',xlab='',ylab=pr)
      drw[,3,] |> plot(type='l',xlab='',ylab=pr)
      drw[,4,] |> plot(type='l',xlab='',ylab=pr)
      par(mar=c(3,5,3,3))
    }

    par(mfrow=c(1,2))
    drw |> hist(main=pr,xlab='')
    drw |> density() |> plot(main=pr)    
  }
}
```


# stan basic

## normal distribution

### ex1-0.stan
normal distribution
```
data{
  int N;
  vector[N] y;
}
parameters{
  real m;
  real<lower=0> s;
}
model{
  y~normal(m,s);
}
generated quantities{
  real y1;
  y1=normal_rng(m,s);
}
```

```{r}
mdl=cmdstan_model('./ex1-0.stan')
y=rnorm(10,2,1)
summary(y)
par(mfrow=c(1,2))
hist(y,main='y')
density(y) |> plot(main='y')  

data=list(N=length(y),y=y)

mle=mdl$optimize(data=data)
mle


mcmc=goMCMC(mdl,data)

mcmc
mcmc$metadata()$stan_variables
mcmc$metadata()$model_params
seeMCMC(mcmc)
```


## poisson distribution

```{r}
y=rpois(20,3)
summary(y)
par(mfrow=c(1,2))
hist(y,main='y')
density(y) |> plot(main='y')

data=list(N=length(y),y=y)
```

### ex2-1.stan
in case fit poisson dist. to normal dist.
```
data{
  int N;
  vector[N] y;
}

parameters{
  real<lower=0> m;
  real<lower=0> s;
}

model{
  y~normal(m,s);
}

generated quantities{
  real y1;
  y1=normal_rng(m,s);
}
```

```{r}
# when fitting poisson dist. sample to normal dist.
mdl=cmdstan_model('./ex2-1.stan')

mle=mdl$optimize(data=data)
mle


mcmc=goMCMC(mdl,data)

seeMCMC(mcmc)

y1=mcmc$draws('y1')
par(mfrow=c(1,2))
hist(y1,main='y1')
density(y1) |> plot(main='y1')     
```

### ex2-2.stan
fit to poisson dist.
```
data{
  int N;
  array[N] int y;
}

parameters{
  real<lower=0> l;
}

model{
  y~poisson(l);
}

generated quantities{
  int y1;
  y1=poisson_rng(l);
}
```


```{r}
mdl=cmdstan_model('./ex2-2.stan')

mle=mdl$optimize(data=data)
mle


mcmc=goMCMC(mdl,data)

seeMCMC(mcmc)

y1=mcmc$draws('y1')
par(mfrow=c(1,2))
hist(y1,main='y1')
density(y1) |> plot(main='y1') 
```


## difference test

### ex3.stan
mean difference of 2 normal distributions
```
data{
  int N;
  vector[N] a;
  vector[N] b;
}
parameters{
  real ma;
  real<lower=0> sa;
  real mb;
  real<lower=0> sb;
}
model{
  a~normal(ma,sa);
  b~normal(mb,sb);
}
generated quantities{
  real d;
  d=mb-ma;
}
```


```{r}
n=20
a=rnorm(n,10,1)
b=rnorm(n,11,2)
data=list(N=n,a=a,b=b)

mdl=cmdstan_model('./ex3.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,ch=F)

d=mcmc$draws('d')
d
mean(d>0)
```


## normal regression

### ex4-1.stan
single normal regression
```
data{
  int N;
  vector[N] x;
  vector[N] y;
}

parameters{
  real b0;
  real b1;
  real<lower=0> s;
}

model{
  y~normal(b0+b1*x,s);
}
```


```{r}
# make prediction with R
n=20
x=runif(n,0,100)
y=rnorm(n,x*3+10,10)
par(mfrow=c(1,1))
plot(x,y)

data=list(N=n,x=x,y=y)

mdl=cmdstan_model('./ex4-1.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,ch=F)

b0=mcmc$draws('b0')
b1=mcmc$draws('b1')
b=tibble(b0=c(b0),b1=c(b1)) |> sample_n(100)
x1=runif(nrow(b),min(x),max(x))
xy=tibble(x=x1,y=b$b0+b$b1*x1)
par(mfrow=c(1,1))
plot(xy)
```


### ex4-2.stan
single normal regression, prediction from new explanatory variable x1
```
data{
  int N;
  vector[N] x;
  vector[N] y;
  int N1;
  vector[N1] x1;
}

parameters{
  real b0;
  real b1;
  real<lower=0> s;
}

model{
  y~normal(b0+b1*x,s);
}

generated quantities{
  vector[N1] m;
  vector[N1] y1;
  for(i in 1:N1){
    m[i]=b0+b1*x1[i];
    y1[i]=normal_rng(m[i],s);
  }
}
```


```{r}
# make prediction with stan
n1=10
x1=seq(min(x),max(x),length.out=n1) # new data fpr predcition

data=list(N=n,x=x,y=y,N1=n1,x1=x1)

mdl=cmdstan_model('./ex4-2.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

mcmc$metadata()$stan_variables
seeMCMC(mcmc,'m',ch=F)

m=mcmc$draws('m')
summary(m)
smm=summary(m)

y1=mcmc$draws('y1')
summary(y1)
smy=summary(y1)

xy=tibble(x=x1,m=smm$median,ml5=smm$q5,mu5=smm$q95,yl5=smy$q5,yu5=smy$q95)

par(mfrow=c(1,1))
xlim=c(min(x),max(x))
ylim=c(min(y),max(y))
plot(x,y,
     xlim=xlim,ylim=ylim, xlab='x',ylab='y',col='red')
par(new=T)
plot(xy$x,xy$m,type='l',
     xlim=xlim,ylim=ylim, xlab='',ylab='',col='black')
par(new=T)
plot(xy$x,xy$ml5,type='l',
     xlim=xlim,ylim=ylim, xlab='',ylab='',col='darkgray')
par(new=T)
plot(xy$x,xy$mu5,type='l',
     xlim=xlim,ylim=ylim, xlab='',ylab='',col='darkgray')
par(new=T)
plot(xy$x,xy$yl5,type='l',
     xlim=xlim,ylim=ylim, xlab='',ylab='',col='lightgray')
par(new=T)
plot(xy$x,xy$yu5,type='l',
     xlim=xlim,ylim=ylim, xlab='',ylab='',col='lightgray')

qplot(x,y,col=I('red'))+
  geom_line(aes(x=x,y=m),xy,col='black')+
  geom_line(aes(x=x,y=ml5),xy,col='darkgray')+
  geom_line(aes(x=x,y=mu5),xy,col='darkgray')+
  geom_line(aes(x=x,y=yl5),xy,col='lightgray')+
  geom_line(aes(x=x,y=yu5),xy,col='lightgray')
```


### ex4-3.stan
single normal regression, prediction from original explanatory variable x
```
data{
  int N;
  vector[N] x;
  vector[N] y;
}

parameters{
  real b0;
  real b1;
  real<lower=0> s;
}

model{
  y~normal(b0+b1*x,s);
}

generated quantities{
  vector[N] m;
  vector[N] y1;
  for(i in 1:N){
    m[i]=b0+b1*x[i];
    y1[i]=normal_rng(m[i],s);
  }
}
```


```{r}
data=list(N=n,x=x,y=y)

mdl=cmdstan_model('./ex4-3.stan')

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,'m',ch=F)

y1=mcmc$draws('y1')
smy=summary(y1)

par(mfrow=c(1,1))
plot(y,smy$median,xlab='obs.',ylab='prd.')
abline(0,1)

qplot(y,smy$median,xlab='obs.',ylab='prd.')+geom_abline(intercept=0,slope=1)

par(mfrow=c(1,2))
hist(y-smy$median,xlab='obs.-prd.',main='residual')
density(y-smy$median) |> plot(xlab='obs.-prd.',main='residual')
```



## estimate correlation

### ex4-41.stan
use covariance matrix and multinormal distribution  
```
data{
  int N;
  array[N] vector[2] y;
}

parameters{
  vector[2] m;
  vector<lower=0>[2] s;
  real<lower=-1,upper=1> r;
}

transformed parameters{
  cov_matrix[2] cv;
  cv[1,1]=s[1]^2;
  cv[2,2]=s[2]^2;
  cv[1,2]=r*s[1]*s[2];
  cv[2,1]=r*s[1]*s[2];
}

model{
  y~multi_normal(m,cv);
}
```

### ex4-42.stan
use covariance matrix and wishart distribution
```
data{
  int N;
  matrix[2,2] dp;
}

parameters{
  vector<lower=0>[2] s;
  real<lower=-1,upper=1> r;
}

transformed parameters{
  cov_matrix[2] cv;
  cv[1,1]=s[1]^2;
  cv[2,2]=s[2]^2;
  cv[1,2]=r*s[1]*s[2];
  cv[2,1]=r*s[1]*s[2];
}

model{
  dp~wishart(N-1,cv);
}
```

### ex4-43.stan
use correlation matrix and wishart distribution
```
data{
  int N;
  int M;
  matrix[M,M] dp;
}

parameters{
  vector<lower=0>[M] s;
  corr_matrix[M] r;
}

transformed parameters{
  cov_matrix[M] cv;
  cv=quad_form_diag(r,s);
}

model{
  dp~wishart(N-1,cv);
}
```

ex4-44.stan
use covariance matrix and multinormal distribution
```
data{
  int N;
  int K;
  array[N] vector[K] y;
}
parameters{
  vector[K] m;
  cov_matrix[K] cov;
}
model{
  y~multi_normal(m,cov);
}
```


```{r}
library(MASS)
n=20
y=mvrnorm(n,c(0,0),matrix(c(1,0.7,0.7,1),2),2)
cov(y)
cor(y)
plot(y)
```

```{r}
#estimate covariance
mdl=cmdstan_model('./ex4-41.stan')

data=list(N=n,y=y)

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,ch=F)
```

```{r}
#estimate correlation,covariance
#deviation product sum matirix folows Wishart dist.
#deviation product sum = covariance * n-1 or n

mdl=cmdstan_model('./ex4-42.stan')

data=list(N=n,dp=cov(y)*(n-1))

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,ch=F)
```

```{r}
#estimate correlation,covariance
mdl=cmdstan_model('./ex4-43.stan')

m=2
data=list(N=n,M=m,dp=cov(y)*(n-1))

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,ch=F)
```

```{r}
mdl=cmdstan_model('./ex4-44.stan')

k=2
data=list(N=n,K=k,y=y)

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc,ch=F)
```



# distributions

## normal dist.
```
distribution of mean, it's from central limit theorem
y~N(m,s), y(-Inf,Inf), E[y]=m, V[y]=s^2
```

### ex1-0.stan
```
data{
  int N;
  vector[N] y;
}
parameters{
  real m;
  real<lower=0> s;
}
model{
  y~normal(m,s);
}
generated quantities{
  real y1;
  y1=normal_rng(m,s);
}
```

```{r}
n=20
y=rnorm(n,2,1)
summary(y)
par(mfrow=c(1,2))
hist(y,main='y')
density(y) |> plot(main='y')

data=list(N=n,y=y)

mdl=cmdstan_model('./ex1-0.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc)
```


## Bernoulli dist.
```
The event with probablity p occur y=1, not occur y=0 
y~Ber(p), y{0,1}, E[y]=p, V[y]=p(1-p)
```
### ex1-1.stan
```
data{
  int N;
  array[N] int y;
}
parameters{
  real<lower=0,upper=1> p;
}
model{
  y~bernoulli(p);
}
generated quantities{
  real s;
  s=(p*(1-p))^.5;
}
```

```{r}
n=10
y=sample(0:1,n,replace=T)
summary(y)
par(mfrow=c(1,2))
hist(y,main='y')
density(y) |> plot(main='y')

data=list(N=n,y=y)

mdl=cmdstan_model('./ex1-1.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc)
```


## geometric dist.
```
The event with probablity p occur after y trials(failure) 
y~Ge(p), y[0,Inf), P(Y=y)=p(1−p)^y
E[y]=1/p, V[y]=(1-p)/p^2
```
### ex1-2.stan
```
data{
  int N;
  array[N] int y;
}
parameters{
  real<lower=0,upper=1> p;
}
model{
  for (n in 1:N) {
    target+=log(p)+y[n]*log(1-p);
  }
}
generated quantities{
  real m;
  m=1/p;
  real s;
  s=((1-p)/p^2)^.5;
}
```

```{r}
n=20
y=rgeom(n,0.3)
summary(y)
par(mfrow=c(1,2))
hist(y,main='y')
density(y) |> plot(main='y')
data=list(N=n,y=y)

mdl=cmdstan_model('./ex1-2.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc)
```


## negative binomial dist.
```
The event with probablity p occur a times after y-a trials 
y~NB(a,p), y[0,Inf), P(Y=y)=p^a*(1−p)^(y-a)
E[y]=a*(1-p)/p, V[y]=a*(1-p)/p^2
```
### ex1-3.stan
```
data{
  int N;
  array[N] int y;
}
parameters{
  real<lower=0,upper=1> p;
  real<lower=1> a;
}
model{
  y~neg_binomial(a,p);
}
generated quantities{
  real m;
  m=a*(1-p)/p;
  real s;
  s=(a*(1-p)/p^2)^.5;
  int y1;
  y1=neg_binomial_rng(a,p);
}
```

```{r}
n=50
a=3
y=rnbinom(n,a,0.3)
summary(y)
par(mfrow=c(1,2))
hist(y,main='y')
density(y) |> plot(main='y')
data=list(N=n,y=y)

mdl=cmdstan_model('./ex1-3.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc)

y1=mcmc$draws('y1')
par(mfrow=c(1,2))
hist(y1,main='y1')
density(y1) |> plot(main='y1') 
```


## binomial dist.
```
The event with probablity p occur y times on n trials
y~B(n,p), y{0,1...n}, E[y]=np, V[y]=np(1-p)
```

### ex2-3.stan
```
data{
  int N;
  int n;
  array[N] int y;
}
parameters{
  real<lower=0,upper=0> p;
}
model{
  y~binomial(n,p);
}
generated quantities{
  real m;
  m=n*p;
  real s;
  s=(n*p*(1-p))^.5;
  int y1;
  y1=binomial_rng(n,p);
}
```

```{r}
n=20
n0=10
y=rbinom(n,n0,0.3)
summary(y)
par(mfrow=c(1,2))
hist(y,main='y')
density(y) |> plot(main='y')
data=list(N=n,n=n0,y=y)

mdl=cmdstan_model('./ex2-3.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc)

y1=mcmc$draws('y1')
par(mfrow=c(1,2))
hist(y1,main='y1')
density(y1) |> plot(main='y1') 
```

### ex2-31.stan
trial n is varied from each sample
```
data{
  int N;
  array[N] int n;
  array[N] int y;
}
parameters{
  real<lower=0> p;
}
model{
  y~binomial(n,p);
}
```

```{r}
n=20
n0=floor(runif(n,1,10))
y=rbinom(n,n0,0.3)

data=list(N=n,n=n0,y=y)

mdl=cmdstan_model('./ex2-31.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc)
```


## Poisson dist.
```
The event occur l times in unit range, the event occur y times. 
y~Po(l), y{0,1...Inf}, E[y]=l, V[y]=l
```

### ex1-4.stan
```
data{
  int N;
  array[N] int y;
}
parameters{
  real<lower=0> l;
}
model{
  y~poisson(l);
}
generated quantities{
  real s;
  s=l^.5;
  int y1;
  y1=poisson_rng(l);
}
```

```{r}
n=20
y=rpois(n,3)
summary(y)
par(mfrow=c(1,2))
hist(y,main='y')
density(y) |> plot(main='y')

data=list(N=n,y=y)

mdl=cmdstan_model('./ex1-4.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc)

y1=mcmc$draws('y1')
par(mfrow=c(1,2))
hist(y1,main='y1')
density(y1) |> plot(main='y1') 
```


## exponential dist.
```
The event occur l times in unit range, the event take time y to occur
y~ex(l), y>0, E[y]=1/l, V[y]=1/l^2
```
### ex1-5.stan
```
data{
  int N;
  vector[N] y;
}
parameters{
  real<lower=0> l;
}
model{
  y~exponential(l);
}
generated quantities{
  real m;
  m=1/l;
  real s;
  s=1/l;
  real y1;
  y1=exponential_rng(l);
}
```

```{r}
n=20
y=rexp(n,2)
summary(y)
par(mfrow=c(1,2))
hist(y,main='y')
density(y) |> plot(main='y')

data=list(N=n,y=y)

mdl=cmdstan_model('./ex1-5.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc)

y1=mcmc$draws('y1')
par(mfrow=c(1,2))
hist(y1,main='y1')
density(y1) |> plot(main='y1') 
```


## gamma dist.
```
The event occur l times in unit range, the event take time y to occur a times
y~Ga(a,l), y>0, E[y]=a/l, V[y]=a/l^2
```
### ex1-6.stan
```
data{
  int N;
  vector[N] y;
}
parameters{
  real<lower=0> a;
  real<lower=0> l;
}
model{
  y~gamma(a,l);
}
```

```{r}
n=20
y=rgamma(n,2,1)
summary(y)
par(mfrow=c(1,2))
hist(y,main='y')
density(y) |> plot(main='y')

data=list(N=n,y=y)

mdl=cmdstan_model('./ex1-6.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc)

y1=mcmc$draws('y1')
par(mfrow=c(1,2))
hist(y1,main='y1')
density(y1) |> plot(main='y1') 
```


## log normal dist.
```
logalithm of variable follows normal distribution
log y~N(m,s), y>0
E[y]=exp(m+s^2)
V[y]=exp(2*m+s^2)*(exp(s^2)-1)
```
### ex1-7.stan
```
data{
  int N;
  vector[N]<lower=0> y;
}
parameters{
  real m0;
  real<lower=0> s0;
}
model{
  y~lognormal(m0,s0);
}
```

```{r}
n=30
y=rlnorm(n,0.5,1)
summary(y)
par(mfrow=c(1,2))
hist(y,main='y')
density(y) |> plot(main='y')

data=list(N=n,y=y)

mdl=cmdstan_model('./ex1-7.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc)

y1=mcmc$draws('y1')
par(mfrow=c(1,2))
hist(y1,main='y1')
density(y1) |> plot(main='y1') 
```



## beta dist.
```
use as prior of binomial dist parameter p
event with probabilty p0 occur a-1 times, do not occur b-1 times in a+b-2 trials
p~Be(a,b), p[0,1], p=p0^(a-1)*(1-p0)^(b-1)
E[p]=a/(a+b), V[p]=ab/(a+b)^2/(a+b+1)
```
### ex1-8.stan
```
data{
  int N;
  vector[N] p;
}
parameters{
  real<lower=0> a;
  real<lower=0> b;
}
model{
  p~beta(a,b);
}
```

```{r}
n=20
p=rbeta(n,1,2)
summary(p)
par(mfrow=c(1,2))
hist(p,main='p')
density(p) |> plot(main='p')

data=list(N=n,p=p)

mdl=cmdstan_model('./ex1-8.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc)
```


## Dirichlet dist.
```
use as prior of categorical dist parameter p
p~dir(p0), p[0,1]
```
### ex1-9.stan
```
data {
  int N;
  int K;
  matrix[N, K] p;
}
parameters {
  simplex[K] p0;
}
model {
  for(i in 1:N){
     p[i]~dirichlet(p0);
  }
}
```

```{r}
library(gtools)
n=20
p=rdirichlet(n,c(0.5,0.3,0.2))
summary(p)
boxplot(p)

data=list(N=n,K=ncol(p),p=p)

mdl=cmdstan_model('./ex1-9.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)

seeMCMC(mcmc)
```


