---
title: "stanEx4"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show 
#date: "`r Sys.Date()`"

---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=T,warning=F,message=F)
suppressWarnings(
  suppressMessages(
    suppressPackageStartupMessages({
      library(stats)
      library(MASS)
      library(tidyverse)
    })
  )
)
```

```{r}
options(scipen=100,digits=3)
library(cmdstanr)
options(mc.cores=parallel::detectCores())
options(cmdstanr_max_rows=1000)
library(gridExtra)
```

execute mcmc sampling
```{r}
goMCMC=function(mdl,data,smp=500,wrm=100,th=1){
  mcmc=mdl$sample(
  data=data,
  seed=1,
  chains=4,
  iter_sampling=smp,
  iter_warmup=wrm,
  thin=th,
  refresh=1000
  )
  mcmc
}
```

see mcmc result and parameters
```{r}
seeMCMC=function(mcmc,exc='',ch=T){ # exclude 'exc' parameters from seeing
  print(mcmc)
  prs=mcmc$metadata()$model_params[-1] # reject lp__
  for(pr in prs){
    if(grepl('^y',pr)) next # not show predictive value "y*" information
    if(exc!='' && grepl(paste0('^',exc),pr)) next
    drw=mcmc$draws(pr)
    if(ch){
      par(mfrow=c(4,1),mar=c(1,5,1,1))
      drw[,1,] |> plot(type='l',xlab='',ylab=pr)
      drw[,2,] |> plot(type='l',xlab='',ylab=pr)
      drw[,3,] |> plot(type='l',xlab='',ylab=pr)
      drw[,4,] |> plot(type='l',xlab='',ylab=pr)
      par(mar=c(3,5,3,3))
    }

    par(mfrow=c(1,2))
    drw |> hist(main=pr,xlab='')
    drw |> density() |> plot(main=pr)    
  }
}
```


```{r}
fn=function(mdl,data,smp=500,wrm=100){
mle=mdl$optimize(data=data)
print(mle)

mcmc=goMCMC(mdl,data,smp,wrm)

mcmc$metadata()$stan_variables
seeMCMC(mcmc,'m')


y0=mcmc$draws('y0')
smy0=summary(y0)

grid.arrange(
  qplot(y,smy0$median,xlab='obs.',ylab='prd.')+geom_abline(intercept=0,slope=1))

par(mfrow=c(1,2))
hist(y-smy0$median,xlab='obs.-prd.',main='residual')
density(y-smy0$median) |> plot(xlab='obs.-prd.',main='residual')


m1=mcmc$draws('m1')
smm1=summary(m1)
y1=mcmc$draws('y1')
smy1=summary(y1)

xy=tibble(x=x1,m=smm1$median,ml5=smm1$q5,mu5=smm1$q95,yl5=smy1$q5,yu5=smy1$q95)

qplot(x,y,col=I('red'))+
  geom_line(aes(x=x,y=ml5),xy,col='darkgray')+
  geom_line(aes(x=x,y=mu5),xy,col='darkgray')+
  geom_line(aes(x=x,y=yl5),xy,col='lightgray')+
  geom_line(aes(x=x,y=yu5),xy,col='lightgray')+
  geom_line(aes(x=x,y=m),xy,col='black')
}
```


## explanatory variable obs.x have noise
x~N(x0.sx),y~N(b0+b1*x0,s)  

### ex8-0.stan
normal regression without explanatory variable's noise
```
data{
  int N;
  vector[N] x;
  vector[N] y;
  int N1;
  vector[N1] x1;
}
parameters{
  real b0;
  real b1;
  real<lower=0> s;
}
model{
  y~normal(b0+b1*x,s);
}
generated quantities{
  vector[N] m0;
  vector[N] y0;
  for(i in 1:N){
    m0[i]=b0+b1*x[i];
    y0[i]=normal_rng(m0[i],s);
  }
  vector[N1] m1;
  vector[N1] y1;
  for(i in 1:N1){
    m1[i]=b0+b1*x1[i];
    y1[i]=normal_rng(m1[i],s);
  }
}
```

### ex9.stan
normal regression with explanatory variable's noise
```
data{
  int N;
  vector[N] x;
  vector[N] y;
  int N1;
  vector[N1] x10;
}
parameters{
  real b0;
  real b1;
  real<lower=0> s;
  real<lower=0> sx;
  vector[N] x0;
}  
model{
  for(i in 1:N){
    x[i]~normal(x0[i],sx);
    y[i]~normal(b0+b1*x0[i],s);
  }
}
generated quantities{
  vector[N] m0;
  vector[N] y0;
  for(i in 1:N){
    m0[i]=b0+b1*x0[i];
    y0[i]=normal_rng(m0[i],s);
  }
  vector[N1] m1;
  vector[N1] x1;
  vector[N1] y1;
  for(i in 1:N1){
    x1[i]=normal_rng(x10[i],sx);
    m1[i]=b0+b1*x10[i];
    y1[i]=normal_rng(m1[i],s);
  }
}
```

```{r}
n=20
x0=runif(n,0,20)
x=rnorm(n,x0,2)
y=rnorm(n,x0*2+5,2)
qplot(x,y)
n1=10

#explanatory variable do not has noise
x1=seq(min(x),max(x),length.out=n1) # new data fpr predcition
data=list(N=n,x=x,y=y,N1=n1,x1=x1)

mdl=cmdstan_model('./ex8-0.stan') 
fn(mdl,data)


#explanatory variable has noise
x10=seq(min(x),max(x),length.out=n1) # new data fpr predcition
data=list(N=n,x=x,y=y,N1=n1,x10=x10)

mdl=cmdstan_model('./ex9.stan')
mcmc=goMCMC(mdl,data,wrm=500,smp=1000)

seeMCMC(mcmc,'[m,x]')

y0=mcmc$draws('y0')
smy0=summary(y0)

grid.arrange(
  qplot(y,smy0$median,xlab='obs.',ylab='prd.')+geom_abline(intercept=0,slope=1))

par(mfrow=c(1,2))
hist(y-smy0$median,xlab='obs.-prd.',main='residual')
density(y-smy0$median) |> plot(xlab='obs.-prd.',main='residual')


m1=mcmc$draws('m1')
smm1=summary(m1)
x1=mcmc$draws('x1')
smx1=summary(x1)
y1=mcmc$draws('y1')
smy1=summary(y1)

xy=tibble(x=smx1$median,m=smm1$median,ml5=smm1$q5,mu5=smm1$q95,yl5=smy1$q5,yu5=smy1$q95)

qplot(x,y,col=I('red'))+
  geom_line(aes(x=x,y=ml5),xy,col='darkgray')+
  geom_line(aes(x=x,y=mu5),xy,col='darkgray')+
  geom_line(aes(x=x,y=yl5),xy,col='lightgray')+
  geom_line(aes(x=x,y=yu5),xy,col='lightgray')+
  geom_line(aes(x=x,y=m),xy,col='black')
```



## outlier

### ex10.stan
objective variable have outlier, y~cauchy(b0+b1*x,s)
```
data{
  int N;
  vector[N] x;
  vector[N] y;
  int N1;
  vector[N1] x1;
}
parameters{
  real b0;
  real b1;
  real<lower=0> s;
}
model{
  y~cauchy(b0+b1*x,s);
}
generated quantities{
  vector[N] m0;
  vector[N] y0;
  for(i in 1:N){
    m0[i]=b0+b1*x[i];
    y0[i]=cauchy_rng(m0[i],s);
  }
  vector[N1] m1;
  vector[N1] y1;
  for(i in 1:N1){
    m1[i]=b0+b1*x1[i];
    y1[i]=cauchy_rng(m1[i],s);
  }
}
```

```{r}
n=20
x=runif(n,0,9)
y=rnorm(n,x*2+5,1)
x[1]=3
y[1]=25
qplot(x,y)
n1=10

x1=seq(min(x),max(x),length.out=n1) # new data fpr predcition
data=list(N=n,x=x,y=y,N1=n1,x1=x1)

mdl=cmdstan_model('./ex8-0.stan') 
fn(mdl,data)

mdl=cmdstan_model('./ex10.stan') 
fn(mdl,data)
```



## censored data

#### objective variable has NA
```
(x,y) i=1-N
(x0,y0) i=1-N0
x1 i=1-N1, y1=NA
(x,y)~N((mx,my),(sx2,sy2,sxy))
(x0,y0)~N((mx,my),(sx2,sy2,sxy))
x1~N(mx,sx2)
```

### ex11-0.stan
```
data{
  int N0;
  array[N0] vector[2] xy;
  int N1;
  vector[N1] x1;
}
parameters{
  vector[2] m;
  cov_matrix[2] s;
}
model{
  target+=multi_normal_lpdf(xy | m, s);
  x1~normal(m[1],s[1,1]^.5);
}
generated quantities{
  vector[2] xy1;
  xy1=multi_normal_rng(m,s);
  real cr;
  cr=s[1,2]/(s[1,1]*s[2,2])^.5;
}
```

```{r}
n=30
x=runif(n,0,9)
y=rnorm(n,10+3*x,4)
cor(x,y)
qplot(x,y)

L=4
n0=sum(x>L)
x0=x[x>L]
y0=y[x>L]
x1=x[x<=L]
n1=sum(x<=L)
cor(x0,y0)
qplot(x0,y0)
```

```{r}
mdl=cmdstan_model('./ex11-0.stan') 

data=list(N0=n0,xy=matrix(c(x0,y0),ncol=2),N1=n1,x1=x1)

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)
seeMCMC(mcmc,ch=F)

xy=mcmc$draws('xy1')
cor(xy[,,1],xy[,,2])
qplot(xy[,,1],xy[,,2])
```


#### objective variable is censored
```
y i=1-N, y~N(m,s)  
  actual          ya i=1-Na
  lower censored  yl i=1-Nl, y<L, P(y<L)=cdf(L-m /s)
  upper censored  yu i=1-Nu, y>U, P(y>U)=ccdf(U-m /s)

cdf(z) cumulative normal density function, P((-Infinity,z],z~N(0,1))
ccdf(z) complementary CDF, P([z,Infinity),z~N(0,1))

P(y | x,m,s)=P(ya i=1-Na)* P(yl i=1-Nl)* P(yu i=1-Nu)
```

### ex11-1.stan
```
data{
  int N;
  vector[N] x;
  vector[N] y;
  real L;
  int Nl;
  vector[Nl] xl;
  real U;
  int Nu;
  vector[Nu] xu;
  int N1;
  vector[N1] x1;
}
parameters{
  real b0;
  real b1;
  real<lower=0> s;
}
model{
  y~normal(b0+b1*x,s);
  for(i in 1:Nl)
    target+=normal_lcdf(L | b0+b1*xl[i],s);
  for(i in 1:Nu)
    target+=normal_lccdf(U | b0+b1*xu[i],s);
}
generated quantities{
  vector[N] m0;
  vector[N] y0;
  for(i in 1:N){
    m0[i]=b0+b1*x[i];
    y0[i]=normal_rng(m0[i],s);
  }
  vector[N1] m1;
  vector[N1] y1;
  for(i in 1:N1){
    m1[i]=b0+b1*x1[i];
    y1[i]=normal_rng(m1[i],s);
  }
}
```

```{r}
n0=20
x=runif(n0,0,9)
y=rnorm(n0,10+3*x,4)
L=15
y[y<L]=L
nl=sum(y==L)
U=30
y[y>U]=U
nu=sum(y==U)
n=n0-nl-nu
qplot(x,y)

xy0=tibble(x=x,y=y)
xya=filter(xy0, y>L & y<U)
xyl=filter(xy0, y==L)
xyu=filter(xy0, y==U)

n1=10
x1=seq(min(x),max(x),length.out=n1) # new data fpr predcition
```


```{r}
mdl=cmdstan_model('./ex8-0.stan')

data=list(N=n,x=xya$x,y=xya$y,N1=n1,x1=x1)

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)
seeMCMC(mcmc,'m')

y0=mcmc$draws('y0')
smy0=summary(y0)

grid.arrange(
  qplot(xya$y,smy0$median,xlab='obs.',ylab='prd.')+geom_abline(intercept=0,slope=1))

par(mfrow=c(1,2))
hist(xya$y-smy0$median,xlab='obs.-prd.',main='residual')
density(xya$y-smy0$median) |> plot(xlab='obs.-prd.',main='residual')

m1=mcmc$draws('m1')
smm1=summary(m1)
y1=mcmc$draws('y1')
smy1=summary(y1)

xy=tibble(x=x1,m=smm1$median,ml5=smm1$q5,mu5=smm1$q95,yl5=smy1$q5,yu5=smy1$q95)

qplot(x,y,col=I('red'))+
  geom_line(aes(x=x,y=ml5),xy,col='darkgray')+
  geom_line(aes(x=x,y=mu5),xy,col='darkgray')+
  geom_line(aes(x=x,y=yl5),xy,col='lightgray')+
  geom_line(aes(x=x,y=yu5),xy,col='lightgray')+
  geom_line(aes(x=x,y=m),xy,col='black')
```


```{r}
data=list(N=n,x=xya$x,y=xya$y,
          L=L,Nl=nl,xl=xyl$x,
          U=U,Nu=nu,xu=xyu$x,
          N1=n1,x1=x1)
mdl=cmdstan_model('./ex11-1.stan') 

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)
seeMCMC(mcmc,'m',ch=T)

y0=mcmc$draws('y0')
smy0=summary(y0)

grid.arrange(
  qplot(xya$y,smy0$median,xlab='obs.',ylab='prd.')+geom_abline(intercept=0,slope=1))

par(mfrow=c(1,2))
hist(xya$y-smy0$median,xlab='obs.-prd.',main='residual')
density(xya$y-smy0$median) |> plot(xlab='obs.-prd.',main='residual')

m1=mcmc$draws('m1')
smm1=summary(m1)
y1=mcmc$draws('y1')
smy1=summary(y1)

xy=tibble(x=x1,m=smm1$median,ml5=smm1$q5,mu5=smm1$q95,yl5=smy1$q5,yu5=smy1$q95)

qplot(x,y,col=I('red'))+
  geom_line(aes(x=x,y=ml5),xy,col='darkgray')+
  geom_line(aes(x=x,y=mu5),xy,col='darkgray')+
  geom_line(aes(x=x,y=yl5),xy,col='lightgray')+
  geom_line(aes(x=x,y=yu5),xy,col='lightgray')+
  geom_line(aes(x=x,y=m),xy,col='black')
```



## sensitivity/specificity analysis

### ex14.stan
estimating sens and spec
```
data {
  int N;
  array[N] int x;
  array[N] int y;
}
parameters {
  real<lower=0,upper=1> p;
  real<lower=0,upper=1> se;
  real<lower=0,upper=1> sp;
}
model {
  p~uniform(0,1);
  se~uniform(0,1);
  sp~uniform(0,1);
  for (i in 1:N) {
    y[i]~bernoulli(x[i]*se+(1-x[i])*(1-sp));
  }
}
generated quantities {
  real ppv;
  real npv;
  ppv=se*p/((se*p)+((1-p)*(1-sp)));
  npv=(1-p)*sp/(((1-p)*sp)+(p*(1-se)));
}
```

```{r}
n=20
data=list(N=n,
          x=sample(0:1,n,replace=T),
          y=sample(0:1,n,replace=T))

mdl=cmdstan_model('./ex14.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)
seeMCMC(mcmc,ch=F)
```



## 2x2 cross table 
```
Effect occur y=1 with probabilty p01, p11 from each cause x{0,1}
event frequncy nxy of effect y{0,1} by cause x{0,1}
n01~B(n0.,p01)
n11~B(n1.,p11)
```

### ex16-1.stan
```
data {
  int n0;
  int n01;
  int n1;
  int n11;
}
parameters {
  real<lower=0,upper=1> p01;
  real<lower=0,upper=1> p11;
}
model {
  n01~binomial(n0,p01);
  n11~binomial(n1,p11);
}
generated quantities {
  real RR;
  RR=p11/p01;
  real OR;
  OR=(p11/(1-p11))/(p01/(1-p01));
}
```

```{r}
n0=20
n01=rbinom(1,n0,0.3)
n1=20
n11=rbinom(1,n1,0.6)
data=list(n0=n0,n01=n01,n1=n1,n11=n11)

mdl=cmdstan_model('./ex16-1.stan')

mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data)
seeMCMC(mcmc,ch=F)
```




## bimodal distribution, mixed normal distribution

```{r}
n=100
y0=rnorm(n,0,1)
y1=rnorm(n,-5,1)
y2=rnorm(n,5,1)
y=sample(c(y0,y1,y2),n)
density(y) |> plot()
```

EM algorithm
```{r}
library(mclust)

rst=Mclust(y)
summary(rst)
rst$parameters
plot(rst)
```

### ex17-1.stan
```
data {
  int N;
  vector[N] y;;
}
parameters {
  simplex[3] h; //ratio of mix
  real m1;
  real m2;
  real m3;
  real<lower=0> s1;
  real<lower=0> s2;
  real<lower=0> s3;
}
model {
  s1~cauchy(0,5);
  s2~cauchy(0,5);
  s3~cauchy(0,5);
  for (i in 1:N) {
    vector[3] p;
    p[1]=log(h[1]) + normal_lpdf(y[i] | m1, s1);
    p[2]=log(h[2]) + normal_lpdf(y[i] | m2, s2);
    p[3]=log(h[3]) + normal_lpdf(y[i] | m3, s3);
    target+=log_sum_exp(p);
  }
}
```

```{r}
mdl=cmdstan_model('./ex17-1.stan')

data=list(N=n,y=y)
mle=mdl$optimize(data=data)
mle

mcmc=goMCMC(mdl,data,smp=2000)
seeMCMC(mcmc,ch=F)
```


