---
title: "Toyota's bayesian analysis text"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide 
#date: "`r Sys.Date()`"

---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=T,warning=F,message=F,comment='')

suppressWarnings(
  suppressMessages(
    suppressPackageStartupMessages({
      library(stats)
      library(MASS)
      library(tidyverse)
      library(magrittr)
      
      library(cmdstanr)
      library(palmerpenguins)　#サンプルデータ用
    })
  )
)
options(scipen=100,digits=3)

options(mc.cores=parallel::detectCores())
options(cmdstanr_max_rows=50)

mode=function(x) {
  as.numeric(names(which.max(table(x))))
}
```

execute mcmc sampling
```{r}
goMCMC=function(mdl,data,smp=500,wrm=100,th=1){
  mcmc=mdl$sample(
  data=data,
  seed=1,
  chains=4,
  iter_sampling=smp,
  iter_warmup=wrm,
  thin=th,
  refresh=1000
  )
  mcmc
}
```

see mcmc result and parameters
```{r}
seeMCMC=function(mcmc,exc='',ch=T){ # exclude 'exc' parameters from seeing
  print(mcmc)
  prs=mcmc$metadata()$model_params[-1] # reject lp__
  for(pr in prs){
    if(grepl('^y',pr)) next # not show predictive value "y*" information
    if(exc!='' && grepl(paste0('^',exc),pr)) next
    drw=mcmc$draws(pr)
    if(ch){
      par(mfrow=c(1,4))
      drw[,1,] |> plot(type='l',main='chain1',ylab=pr)
      drw[,2,] |> plot(type='l',main='chain2',ylab=pr)
      drw[,3,] |> plot(type='l',main='chain3',ylab=pr)
      drw[,4,] |> plot(type='l',main='chain4',ylab=pr)
    }

    par(mfrow=c(1,2))
    drw |> hist(main=pr,xlab='')
    drw |> density() |> plot(main=pr)    
  }
}
```


## estimate parameters

stan01.stan
```
data {
  int<lower=0> N; //number of observation
  vector[N] y; //observations
}
parameters {
  real m;//mean
  real<lower=0> s; //standard deviation
}
model {
  y~normal(m,s); //data genaration
}
generated quantities {
  real y1; //data prediction
  y1 = normal_rng(m, s);
  real ll; //log liklihood
  ll = normal_lpdf(y|m,s);
}
```

```{r}
#「牛丼データ」の入力
y=c(76.5,83.9,87.9,70.8,84.6,85.1,79.6,79.8,79.7,78.0)
data=list(N=length(y),y=y)

mdl=cmdstan_model('./stan01.stan')

mle=mdl$optimize(data=data)
mle

system.time({
  mcmc=goMCMC(mdl,data)
  seeMCMC(mcmc,ch=F)
})

c=85

m=mcmc$draws('m') |> as.vector()
mode(round(m))
quantile(m,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
sum(m<c)/length(m)

s=mcmc$draws('s') |> as.vector()
mode(round(s))
quantile(s,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

quantile(s/m,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

mode(round((m-c)/s))
quantile((m-c)/s,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

y1=mcmc$draws('y1') |> as.vector()
density(y1) |> plot(main='y1')
mode(round(y1))
quantile(y1,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

sum(y1<c)/length(y1)

quantile(y1/c,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
```


```{r}
### 第1章章末問題
y=c(
36,38,51,40,41,52,43,31,35,37,49,43,43,41,36,53,43,26,45,37,
33,38,33,35,36,28,46,41,32,49,43,38,46,46,46,45,44,40,38,37,
35,39,31,55,48,32,37,37,45,39,42,40,40,50,38,51,29,44,41,42,
43,36,38,33,32,42,43,40,46,54,37,24,47,35,35,47,38,31,41,39,
40,43,37,45,38,42,48,43,38,48,47,44,42,36,50,36,55,51,38,33)
data=list(N=length(y),y=y)

mdl=cmdstan_model('./stan01.stan')

mle=mdl$optimize(data=data)
mle

system.time({
  mcmc=goMCMC(mdl,data)
  seeMCMC(mcmc,ch=F)
})

c=50

m=mcmc$draws('m') |> as.vector()
mode(round(m))
quantile(m,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
sum(m<c)/length(m)

s=mcmc$draws('s') |> as.vector()
mode(round(s))
quantile(s,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

quantile(s/m,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

mode(round((m-c)/s))
quantile((m-c)/s,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

y1=mcmc$draws('y1') |> as.vector()
density(y1) |> plot(main='y1')
mode(round(y1))
quantile(y1,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

sum(y1<c)/length(y1)

quantile(y1/c,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

```


## independent 2 groups
```{r}
#表3.1のデータ、クラスA、クラスBの順に入力
y1=c(49,66,69,55,54,72,51,76,40,62,66,51,59,68,66,57,53,66,58,57)
y2=c(41,55,21,49,53,50,52,67,54,69,57,48,31,52,56,50,46,38,62,59)
data=list(N=length(y1),y=y1)

mdl=cmdstan_model('./stan01.stan')

mle=mdl$optimize(data=data)
mle

system.time({
  mcmc1=goMCMC(mdl,data)
  seeMCMC(mcmc1,ch=F)
})

m1=mcmc1$draws('m') |> as.vector()
s1=mcmc1$draws('s') |> as.vector()
y11=mcmc1$draws('y1') |> as.vector()


mdl=cmdstan_model('./stan01.stan')
data=list(N=length(y2),y=y2)

mle=mdl$optimize(data=data)
mle

system.time({
  mcmc2=goMCMC(mdl,data)
  seeMCMC(mcmc2,ch=F)
})

m2=mcmc2$draws('m') |> as.vector()
s2=mcmc2$draws('s') |> as.vector()
y12=mcmc2$draws('y1') |> as.vector()


quantile(m1-m2,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

sum(m1>m2)/length(m1)
sum(m1-m2>3)/length(m1)
sum(m1-m2>5)/length(m1)
sum(m1-m2>10)/length(m1)

quantile((m1-m2)/s1,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
c=0.3
sum((m1-m2)/s1>c)/length(m1)

u=pnorm(m1,m2,s2) ##cohen's U3, third measure of non-overlap
quantile(u,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

pd=pnorm((m1-m2)/s1/2^.5,0,1)
quantile(pd,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1)) #probability of dominance

c=3
pc=pnorm((m1-m2-c)/s1/2^.5,0,1)
quantile(pc,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1)) #probability beyond threshold


density(y11) |> plot(main='y1')
mode(round(y11))
quantile(y11,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

density(y12) |> plot(main='y2')
mode(round(y12))
quantile(y12,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
```


```{r}
#データ入力　罹患群を第1群，健常群を第2群とする
y1=c(
56,55,55,62,54,63,47,58,56,56,57,52,53,50,50,57,57,55,60,65,53,43,60,51,52,
60,54,49,56,54,55,57,53,58,54,57,60,57,53,61,60,58,56,52,62,52,66,63,54,50)
y2=c(
33,37,59,41,42,61,46,25,32,35,55,44,45,41,33,61,46,16,48,34,27,37,28,31,32,
20,50,42,26,55,45,36,51,51,50,48,47,39,36,35,32,38,25,66,54,27,35,34,49,39)
data=list(N=length(y1),y=y1)

mdl=cmdstan_model('./stan01.stan')

mle=mdl$optimize(data=data)
mle

system.time({
  mcmc1=goMCMC(mdl,data)
  seeMCMC(mcmc1,ch=F)
})

m1=mcmc1$draws('m') |> as.vector()
s1=mcmc1$draws('s') |> as.vector()
y11=mcmc1$draws('y1') |> as.vector()


mdl=cmdstan_model('./stan01.stan')
data=list(N=length(y2),y=y2)

mle=mdl$optimize(data=data)
mle

system.time({
  mcmc2=goMCMC(mdl,data)
  seeMCMC(mcmc2,ch=F)
})

m2=mcmc2$draws('m') |> as.vector()
s2=mcmc2$draws('s') |> as.vector()
y12=mcmc2$draws('y1') |> as.vector()


quantile(m1-m2,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
c=15
sum(m1-m2>c)/length(m1)

quantile((m1-m2)/s1,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
c=3
sum((m1-m2)/s1>c)/length(m1)

u=pnorm(m1,m2,s2) ##cohen's U3, third measure of non-overlap
quantile(u,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

pd=pnorm((m1-m2)/s1/2^.5,0,1)
quantile(pd,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1)) #probability of dominance

c=15
pc=pnorm((m1-m2-c)/s1/2^.5,0,1)
quantile(pc,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1)) #probability beyond threshold


density(y11) |> plot(main='y1')
mode(round(y11))
quantile(y11,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

density(y12) |> plot(main='y2')
mode(round(y12))
quantile(y12,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
```



## corresponding 2 groups, correlation

stan02.stan
```
data{
  int N;
  int K;
  array[N] vector[K] y;
}
parameters{
  vector[K] m;
  cov_matrix[K] cv;
}
model{
  y~multi_normal(m,cv);
}
generated quantities{
  vector[K] y1;
  y1=multi_normal_rng(m,cv);
  real ll;
  ll=multi_normal_lpdf(y|m,cv);
}
```  

```{r}
#表4.1の「ダイエット」データ入力
y1=c(53.1,51.5,45.5,55.5,49.6,50.1,59.2,54.7,53.0,48.6,55.3,52.6,51.7,
      48.6,56.4,42.9,50.3,42.4,51.2,39.1)
y2=c(48.3,45.2,46.6,56.6,41.2,44.6,51.9,55.5,45.4,47.6,50.6,54.5,49.0,
      43.9,53.8,40.1,52.8,35.3,55.6,38.0)
y=cbind(y1,y2)
data=list(N=nrow(y),K=ncol(y),y=y)

mdl=cmdstan_model('./stan02.stan')

mle=mdl$optimize(data=data)
mle

system.time({
  mcmc=goMCMC(mdl,data)
  seeMCMC(mcmc,ch=F)
})

m1=mcmc$draws('m[1]') |> as.vector()
m2=mcmc$draws('m[2]') |> as.vector()

s1=mcmc$draws('cv[1,1]') |> as.vector() |> sqrt()
s2=mcmc$draws('cv[2,2]') |> as.vector() |> sqrt()
cv=mcmc$draws('cv[1,2]') |> as.vector()
r=cv/s1/s2

y11=mcmc$draws('y1[1]') |> as.vector()
y12=mcmc$draws('y1[2]') |> as.vector()


sum(m1>m2)/length(m1)
quantile(m1-m2,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
c=2
sum(m1-m2>c)/length(m1)

quantile((m1-m2)/s1,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
c=0.3
sum((m1-m2)/s1>c)/length(m1)

u=pnorm(m1,m2,s2) ##cohen's U3, third measure of non-overlap
quantile(u,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
c=0.6
sum(u>c)/length(u)

pd=pnorm((m1-m2)/s1/2^.5,0,1)
quantile(pd,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1)) #probability of dominance

c=1
pc=pnorm((m1-m2-c)/s1/2^.5,0,1)
quantile(pc,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1)) #probability beyond threshold


density(y11) |> plot(main='y1')
mode(round(y11))
quantile(y11,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

density(y12) |> plot(main='y2')
mode(round(y12))
quantile(y12,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))


d=y1-y2
quantile(d,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

s12=(s1^2+s2^2-2*r*s1*s2)^.5
quantile(s12,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
c=3
sum(s12<c)/length(s12)

quantile((m1-m2)/s12,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
c=0.3
sum((m1-m2)/s12>0.3)/length(m1)

pd=pnorm((m1-m2)/s12,0,1)
quantile(pd,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1)) #probability of dominance
c=0.8
sum(pd>c)/length(pd)

c=3
pc=pnorm((m1-m2-c)/s12,0,1)
quantile(pc,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1)) #probability beyond threshold

quantile(r,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
sum(r>0.6 & r<0.8)/length(r)

pco=0.5+asin(r)/pi
quantile(pco,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1)) #probability of concordance
sum(pco>0.8)/length(pco)
```


```{r}
#データ 援助後を第1群，援助前を第2群とする
y1=c(
73,72,56,58,71,42,78,77,75,72,56,71,69,77,84,51,62,88,56,58,84,91,71,82,81,
77,65,78,79,60,66,70,65,57,64,61,56,67,75,64,68,67,80,55,48,85,56,62,65,79)
y2=c(
62,54,19,54,47,22,35,77,64,60,27,41,41,44,57,16,42,89,40,67,69,46,74,62,60,
87,32,42,73,25,42,57,31,35,33,38,43,53,55,62,67,56,76,05,31,70,66,65,34,48)
y=cbind(y1,y2)
data=list(N=nrow(y),K=ncol(y),y=y)

mdl=cmdstan_model('./stan02.stan')

mle=mdl$optimize(data=data)
mle

system.time({
  mcmc=goMCMC(mdl,data)
  seeMCMC(mcmc,ch=F)
})

m1=mcmc$draws('m[1]') |> as.vector()
m2=mcmc$draws('m[2]') |> as.vector()

s1=mcmc$draws('cv[1,1]') |> as.vector() |> sqrt()
s2=mcmc$draws('cv[2,2]') |> as.vector() |> sqrt()
cv=mcmc$draws('cv[1,2]') |> as.vector()
r=cv/s1/s2

y11=mcmc$draws('y1[1]') |> as.vector()
y12=mcmc$draws('y1[2]') |> as.vector()


sum(m1>m2)/length(m1)
quantile(m1-m2,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
c=20
sum(m1-m2>c)/length(m1)

quantile((m1-m2)/s1,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
c=1.5
sum((m1-m2)/s1>c)/length(m1)

u=pnorm(m1,m2,s2) ##cohen's U3, third measure of non-overlap
quantile(u,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
c=0.8
sum(u>c)/length(u)

pd=pnorm((m1-m2)/s1/2^.5,0,1)
quantile(pd,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1)) #probability of dominance

c=1
pc=pnorm((m1-m2-c)/s1/2^.5,0,1)
quantile(pc,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1)) #probability beyond threshold


density(y11) |> plot(main='y1')
mode(round(y11))
quantile(y11,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

density(y12) |> plot(main='y2')
mode(round(y12))
quantile(y12,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))


d=y1-y2
quantile(d,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

s12=(s1^2+s2^2-2*r*s1*s2)^.5
quantile(s12,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
c=3
sum(s12<c)/length(s12)

quantile((m1-m2)/s12,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
c=0.3
sum((m1-m2)/s12>0.3)/length(m1)

pd=pnorm((m1-m2)/s12,0,1)
quantile(pd,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1)) #probability of dominance
c=0.8
sum(pd>c)/length(pd)

c=20
pc=pnorm((m1-m2-c)/s12,0,1)
quantile(pc,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1)) #probability beyond threshold

quantile(r,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
sum(r>0.6)/length(r)

pco=0.5+asin(r)/pi
quantile(pco,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1)) #probability of concordance
sum(pco>0.7)/length(pco)
```



## independent one factorial design

stan03.stan
```
data{
  int N;
  int K;
  vector[N] y;
  array[N] int a;
}
parameters{
  vector[K] m;
  real<lower=0> s;
}
model{
  y~normal(m[a],s);
}
generated quantities{
  real ll;
  ll=normal_lpdf(y | m[a], s);
}
```

```{r}
#表5.1の「亜硫酸ガスデータ」の入力
k=4
a=rep(1:k,each=6);
y=c(
  10, 10, 09, 11, 12, 11,   08, 10, 08, 10, 12, 09,
  08, 08, 11, 11, 14, 15,   14, 12, 11, 16, 13, 12)
data=list(N=length(y),K=k,y=y,a=a)

mdl=cmdstan_model('./stan03.stan')

mle=mdl$optimize(data=data)
mle

system.time({
  mcmc=goMCMC(mdl,data)
  seeMCMC(mcmc,ch=F)
})


m=tibble(m1=mcmc$draws('m[1]') |> as.vector(),
         m2=mcmc$draws('m[2]') |> as.vector(),
         m3=mcmc$draws('m[3]') |> as.vector(),
         m4=mcmc$draws('m[4]') |> as.vector()) |> as.matrix()
s=mcmc$draws('s') |> as.vector()

m0=apply(m,1,mean)
quantile(m0,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

a=m-m0
for(i in 1:k){
  quantile(a[,k],probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
}

sa=(apply(a^2,1,sum)/k)^.5
quantile(sa,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

eta2=sa^2/(sa^2+s^2) #coefficient of determination
quantile(eta2,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

dlt=sa/s #effect size
quantile(dlt,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

for(i in 1:(k-1)){
  for(j in (i+1):k){
    cat('\n',i,' > ',j,sum(m[,i]>m[,j])/nrow(m))
  }
}
```


```{r}
#表5.15 マウスの体重のデータ
k=3
a=c(rep(1,8),rep(2,9),rep(3,10))
y=c(05.02, 06.67, 08.17, 02.79, 08.13, 06.34, 06.32, 03.97,
    09.89, 09.58, 11.20, 09.05, 12.33, 09.39, 10.88, 09.37, 17.40,
    10.20, 07.29, 07.57, 03.42, 05.82, 10.92, 05.21, 13.47, 08.64, 06.05)
data=list(N=length(y),K=k,y=y,a=a)

mdl=cmdstan_model('./stan03.stan')

mle=mdl$optimize(data=data)
mle

system.time({
  mcmc=goMCMC(mdl,data)
  seeMCMC(mcmc,ch=F)
})

m=tibble(m1=mcmc$draws('m[1]') |> as.vector(),
         m2=mcmc$draws('m[2]') |> as.vector(),
         m3=mcmc$draws('m[3]') |> as.vector()) |> as.matrix()
s=mcmc$draws('s') |> as.vector()

m0=apply(m,1,mean)
quantile(m0,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

a=m-m0
for(i in 1:k){
  quantile(a[,k],probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
}

sa=(apply(a^2,1,sum)/k)^.5
quantile(sa,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

eta2=sa^2/(sa^2+s^2) #coefficient of determination
quantile(eta2,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

dlt=sa/s #effect size
quantile(dlt,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

for(i in 1:(k-1)){
  for(j in (i+1):k){
    cat('\n',i,' > ',j,sum(m[,i]>m[,j])/nrow(m))
  }
}
```


## independent two factorial design

stan04.stan
```
data {
  int N;
  int Ka;
  int Kb;
  vector[N] y;
  array[N] int a;
  array[N] int b;
}

parameters {
  real m;
  vector[Ka-1] ma0;
  vector[Kb-1] mb0;
  matrix[Ka-1, Kb-1] mab0;
  real<lower=0> s;
}

transformed parameters {
  vector[Ka] ma;
  vector[Kb] mb;
  matrix[Ka, Kb] mab;

// constrain sum(ma)=0
  for (i in 1:(Ka-1)) {ma[i]=ma0[i];}
  ma[Ka]=-sum(ma0);

// constrain sum(mb)=0
  for (i in 1:(Kb-1)) {mb[i]=mb0[i];}
  mb[Kb]=-sum(mb0);

// constrain sum(mab)=0
  for (i in 1:(Ka-1)) {
    for (j in 1:(Kb-1)) {mab[i, j]=mab0[i, j];}
  }

  // constrain sum(mab's each rows)
  for (i in 1:(Ka-1)) {mab[i, Kb]=-sum(mab0[i, 1:(Kb-1)]);}

  // constrain sum(mab's each columns)
  for (j in 1:(Kb-1)) {mab[Ka, j]=-sum(mab0[1:(Ka-1), j]);}

  mab[Ka, Kb]=sum(mab[1:(Ka-1), 1:(Kb-1)]);
}

model {
  for (i in 1:N) {y[i]~normal(m+ma[a[i]]+mb[b[i]]+mab[a[i], b[i]], s);}
}

generated quantities {
  real ll=0.0;
  for (i in 1:N) {ll+=normal_lpdf(y[i] | m+ma[a[i]]+mb[b[i]]+mab[a[i], b[i]], s);}
}
```

```{r}
#表5.8の「サンフランシスコとロサンゼルスのホテルの料金」の入力
ka=2
kb=2
a=rep(1:2,each=18)
b=rep(1:2,each=9,times=2)
y=c(
 079,107,103,092,180,165,240,265,300,
 075,060,060,094,119,100,102,125,165,
 095,099,070,116,170,145,205,200,210,
 153,078,075,092,115,155,250,340,380)
data=list(N=length(y),Ka=ka,Kb=kb,y=y,a=a,b=b)

mdl=cmdstan_model('./stan04.stan')

mle=mdl$optimize(data=data)
mle

system.time({
  mcmc=goMCMC(mdl,data)
  #seeMCMC(mcmc,ch=F)
})

mcmc
mcmc$summary(variables=c('m','s','ma[1]','mb[1]','mab[1,1]','ll'))

m=mcmc$draws('m') |> as.vector()
quantile(m,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

ma=tibble(ma1=mcmc$draws('ma[1]') |> as.vector(),
          ma2=mcmc$draws('ma[2]') |> as.vector()) |> as.matrix()
mb=tibble(mb1=mcmc$draws('mb[1]') |> as.vector(),
          mb2=mcmc$draws('mb[2]') |> as.vector()) |> as.matrix()
mab=tibble(mab11=mcmc$draws('mab[1,1]') |> as.vector(),
           mab12=mcmc$draws('mab[1,2]') |> as.vector(),
           mab21=mcmc$draws('mab[2,1]') |> as.vector(),
           mab22=mcmc$draws('mab[2,2]') |> as.vector()) |> as.matrix()

quantile(m+ma[,1]+mb[,1]+mab[,'mab11'],probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(m+ma[,1]+mb[,2]+mab[,'mab12'],probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(m+ma[,2]+mb[,1]+mab[,'mab21'],probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(m+ma[,2]+mb[,2]+mab[,'mab22'],probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

s=mcmc$draws('s') |> as.vector()

sa=(apply(ma^2,1,sum)/ka)^.5
sb=(apply(mb^2,1,sum)/kb)^.5
sab=(apply(mab^2,1,sum)/ka/kb)^.5
sy2=sa^2+sb^2+sab^2+s^2

eta2a=sa^2/sy2 #coefficient of determination
quantile(eta2a,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

eta2b=sb^2/sy2 #coefficient of determination
quantile(eta2b,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

eta2ab=sab^2/sy2 #coefficient of determination
quantile(eta2ab,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

eta2=(sa^2+sb^2+sab^2)/sy2 #coefficient of determination
quantile(eta2,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

dlt_a=sa/s #effect size
quantile(dlt_a,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

dlt_b=sb/s #effect size
quantile(dlt_b,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

dlt_ab=sab/s #effect size
quantile(dlt_ab,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
```


```{r}
#表5.16 走者の有無による選手Ｅの球種別の球速
a=c(rep(1,49),rep(2,49))
b=c(rep(1,10),rep(2,8),rep(3,7),rep(4,9),rep(5,8),rep(6,7),
     rep(1,10),rep(2,8),rep(3,7),rep(4,9),rep(5,8),rep(6,7))
y=c(140,146,149,136,147,147,143,143,143,141,
139,136,136,140,135,132,140,134,
123,127,131,130,138,128,129,
115,120,118,118,121,124,129,119,128,
128,124,123,121,122,126,131,122,
121,121,120,116,117,113,118,
143,141,142,145,149,145,143,141,142,155,
138,134,142,136,135,136,131,133,
131,128,128,128,127,130,130,
117,125,132,122,119,122,129,117,127,
117,120,124,122,122,122,118,122,
119,125,122,116,119,113,122)
ka=length(unique(a))
kb=length(unique(b))

data=list(N=length(y),Ka=ka,Kb=kb,y=y,a=a,b=b)

mdl=cmdstan_model('./stan04.stan')

mle=mdl$optimize(data=data)
mle

system.time({
  mcmc=goMCMC(mdl,data)
  #seeMCMC(mcmc,ch=F)
})

mcmc

m=mcmc$draws('m') |> as.vector()
quantile(m,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

ma=tibble(ma1=mcmc$draws('ma[1]') |> as.vector(),
          ma2=mcmc$draws('ma[2]') |> as.vector()) |> as.matrix()
mb=tibble(mb1=mcmc$draws('mb[1]') |> as.vector(),
          mb2=mcmc$draws('mb[2]') |> as.vector(),
          mb3=mcmc$draws('mb[3]') |> as.vector(),
          mb4=mcmc$draws('mb[4]') |> as.vector(),
          mb5=mcmc$draws('mb[5]') |> as.vector(),
          mb6=mcmc$draws('mb[6]') |> as.vector()) |> as.matrix()
mab=tibble(mab11=mcmc$draws('mab[1,1]') |> as.vector(),
           mab12=mcmc$draws('mab[1,2]') |> as.vector(),
           mab13=mcmc$draws('mab[1,3]') |> as.vector(),
           mab14=mcmc$draws('mab[1,4]') |> as.vector(),
           mab15=mcmc$draws('mab[1,5]') |> as.vector(),
           mab16=mcmc$draws('mab[1,6]') |> as.vector(),
           mab21=mcmc$draws('mab[2,1]') |> as.vector(),
           mab22=mcmc$draws('mab[2,2]') |> as.vector(),
           mab23=mcmc$draws('mab[2,3]') |> as.vector(),
           mab24=mcmc$draws('mab[2,4]') |> as.vector(),
           mab25=mcmc$draws('mab[2,5]') |> as.vector(),
           mab26=mcmc$draws('mab[2,6]') |> as.vector()) |> as.matrix()

quantile(m+ma[,1]+mb[,1]+mab[,'mab11'],probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(m+ma[,1]+mb[,2]+mab[,'mab12'],probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(m+ma[,1]+mb[,3]+mab[,'mab13'],probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(m+ma[,1]+mb[,4]+mab[,'mab14'],probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(m+ma[,1]+mb[,5]+mab[,'mab15'],probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(m+ma[,1]+mb[,6]+mab[,'mab16'],probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(m+ma[,2]+mb[,1]+mab[,'mab21'],probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(m+ma[,2]+mb[,2]+mab[,'mab22'],probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(m+ma[,2]+mb[,3]+mab[,'mab23'],probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(m+ma[,2]+mb[,4]+mab[,'mab24'],probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(m+ma[,2]+mb[,5]+mab[,'mab25'],probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(m+ma[,2]+mb[,6]+mab[,'mab26'],probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

s=mcmc$draws('s') |> as.vector()

sa=(apply(ma^2,1,sum)/ka)^.5
quantile(sa,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

sb=(apply(mb^2,1,sum)/kb)^.5
quantile(sb,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

sab=(apply(mab^2,1,sum)/ka/kb)^.5
quantile(sab,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

sy2=sa^2+sb^2+sab^2+s^2

eta2a=sa^2/sy2 #coefficient of determination
quantile(eta2a,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

eta2b=sb^2/sy2 #coefficient of determination
quantile(eta2b,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

eta2ab=sab^2/sy2 #coefficient of determination
quantile(eta2ab,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

eta2=(sa^2+sb^2+sab^2)/sy2 #coefficient of determination
quantile(eta2,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

dlt_a=sa/s #effect size
quantile(dlt_a,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

dlt_b=sb/s #effect size
quantile(dlt_b,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

dlt_ab=sab/s #effect size
quantile(dlt_ab,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
```


## discrete quantities
### proportion

stan05-1.stan
```
data { 
  int n; 
  int x; 
}
parameters {
  real<lower=0,upper=1> p;
}
transformed parameters {
}
model {
  x~binomial(n,p);
}
generated quantities{
  int x1;
  x1=binomial_rng(n,p);
  real ll;
  ll=binomial_lpmf(x|n,p);
}
```

```{r}
#6.2 蕎麦の選好問題
n=400
x=220
data=list(n=n,x=x)

mdl=cmdstan_model('./stan05-1.stan')

mle=mdl$optimize(data=data)
mle

system.time({
  mcmc=goMCMC(mdl,data)
  seeMCMC(mcmc,ch=F)
})

p=mcmc$draws('p') |> as.vector()

sum(p>0.5)/length(p)

od=p/(1-p) #odds
quantile(od,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
```

```{r}
#6.4.1 法案賛否問題 1
n1=120
x1=71
data1=list(n=n1,x=x1)
mdl=cmdstan_model('./stan05-1.stan')

mle=mdl$optimize(data=data1)
mle

system.time({
  mcmc1=goMCMC(mdl,data1)
  seeMCMC(mcmc1,ch=F)
})

n2=125
x2=42
data2=list(n=n2,x=x2)
mdl=cmdstan_model('./stan05-1.stan')

mle=mdl$optimize(data=data2)
mle

system.time({
  mcmc2=goMCMC(mdl,data2)
  seeMCMC(mcmc2,ch=F)
})

p1=mcmc1$draws('p') |> as.vector()
sum(p1>0.5)/length(p1)
od1=p1/(1-p1) #odds
quantile(od1,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

p2=mcmc2$draws('p') |> as.vector()
sum(p2>0.5)/length(p2)
od2=p2/(1-p2) #odds
quantile(od2,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

quantile(p1-p2,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(p1/p2,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(od1/od2,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
```


### count data

stan05-2.stan
```
data { 
  int K; 
  array[K] int n; 
}
parameters {
  simplex[K] p;
}
transformed parameters {
}
model {
  n~multinomial(p);
}
generated quantities{
  real ll;
  ll=multinomial_lpmf(n|p);
}
```

```{r}
#6.3 相談相手問題
n=c(26,40,8,2,23,8)
data=list(K=length(n),n=n)

mdl=cmdstan_model('./stan05-2.stan')

mle=mdl$optimize(data=data)
mle

system.time({
  mcmc=goMCMC(mdl,data)
  seeMCMC(mcmc,ch=F)
})
```


### cross table, corresponding discrete variables

stan05-3.stan
```
data { 
  int R;
  int C;
  array[R,C] int n;
  int N;
}
transformed data {
  array[R*C] int nv; 
  for (j in 1:C)
    for (i in 1:R)
      nv[(j-1)*R+i]=n[i,j];
}
parameters {
  simplex[R*C] p; 
}
model {
  nv~multinomial(p); 
}
generated quantities {
  array[R*C] int n1=multinomial_rng(p,N);
  real ll=multinomial_lpmf(nv|p);
}
```

```{r}
#6.5.1 法案賛否問題2
r=2
c=2
n=matrix(c(55,16,14,35),r,c)
data=list(R=r,C=c,n=n,N=sum(n))

mdl=cmdstan_model('./stan05-3.stan')

mle=mdl$optimize(data=data)
mle

system.time({
  mcmc=goMCMC(mdl,data)
  #seeMCMC(mcmc,ch=F)
})

mcmc

p11=mcmc$draws('p[1]') |> as.vector()
p21=mcmc$draws('p[2]') |> as.vector()
p12=mcmc$draws('p[3]') |> as.vector()
p22=mcmc$draws('p[4]') |> as.vector()

p10=p11+p12
p20=p21+p22
p01=p11+p21
p02=p12+p22

e11=(p11-p10*p01)/(p10*p01)^.5 # Peason's residual
e21=(p21-p20*p01)/(p20*p01)^.5
e12=(p12-p10*p02)/(p10*p02)^.5
e22=(p22-p20*p02)/(p20*p02)^.5

quantile(e11,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(e21,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(e12,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(e22,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

v=((e11^2+e21^2+e12^2+e22^2)/(min(r,c)-1))^.5 # Cramer's V
quantile(v,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
```


```{r}
#6.5.4 ワイン選択問題
r=3
c=3
n=matrix(c(19,12,6, 8,8,4, 15,19,18),r,c)
data=list(R=r,C=c,n=n,N=sum(n))

mdl=cmdstan_model('./stan05-3.stan')

mle=mdl$optimize(data=data)
mle

system.time({
  mcmc=goMCMC(mdl,data)
  #seeMCMC(mcmc,ch=F)
})

mcmc

p11=mcmc$draws('p[1]') |> as.vector()
p21=mcmc$draws('p[2]') |> as.vector()
p31=mcmc$draws('p[3]') |> as.vector()
p12=mcmc$draws('p[4]') |> as.vector()
p22=mcmc$draws('p[5]') |> as.vector()
p32=mcmc$draws('p[6]') |> as.vector()
p13=mcmc$draws('p[7]') |> as.vector()
p23=mcmc$draws('p[8]') |> as.vector()
p33=mcmc$draws('p[9]') |> as.vector()

p10=p11+p12+p13
p20=p21+p22+p23
p30=p31+p32+p33
p01=p11+p21+p31
p02=p12+p22+p32
p03=p13+p23+p33

quantile(p10,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(p20,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(p30,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(p01,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(p02,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
quantile(p03,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))

e11=(p11-p10*p01)/(p10*p01)^.5 # Peason's residual
e21=(p21-p20*p01)/(p20*p01)^.5
e31=(p31-p30*p01)/(p30*p01)^.5
e12=(p12-p10*p02)/(p10*p02)^.5
e22=(p22-p20*p02)/(p20*p02)^.5
e32=(p32-p30*p02)/(p30*p02)^.5
e13=(p13-p10*p03)/(p10*p03)^.5
e23=(p23-p20*p03)/(p20*p03)^.5
e33=(p33-p30*p03)/(p30*p03)^.5

v=((e11^2+e21^2+e31^2+e12^2+e22^2+e32^2+e13^2+e23^2+e33^2)/(min(r,c)-1))^.5 # Cramer's V
quantile(v,probs=c(0.0,0.025,0.05,0.25,0.5,0.75,0.95,0.975,1))
```

